HO unification from object language to meta language

### 2024-03-27 paper/main.tex
wip

\newtheorem{invariant}{Invariant}

@@

\begin{invariant}[Link left hand side]\label{inv:linklhs}
  The left hand side of a new link
  is a variable.
\end{invariant}

\noindent
If the variable is assigned during a run the link is considered for
progress and possibly eliminated. This is discussed in
section~\ref{sec:eta}.\marginpar{is this the right section?}

@@

TODO: discuss occur check

@@

\begin{invariant}[\linketa \rhs]
  The \rhs of a \linketa having the shape $\lambda x.F_x$
  where $F_x$ is a term not starting with the \elpiIn{lam} constructor.
  \label{invariant:link-eta-right}
\end{invariant}

In the next three subsections we explain how we detect \maybeeta terms, how we
compile them and how \linketa are activated during the execution of the program
and provide justification for why these two invariants remain true.

@@

As a remark, note that $\lambda x. f \appsep A_x \appsep x$ is a \maybeeta,
since, despite, $x$ occurs in $f \appsep A_x$, it is still possible that this
subterm does not use $x$, for example if $A$ is a function on the form $\lambda
x.a$, where $a$ is a defined constant. In this case, the \maybeeta should
consider that the bound variable \elpiIn{x} does not ``rigidly'' occur in 
the given subterm.

@@

\paragraph{Detection of \maybeeta} The main modification of the compiler to solve this
unification issue consists in identifying all the subterms of a term $t$ that
are \maybeeta. In particular, a term $t$ is a \maybeeta, if it can reduce to a
$\eta$-expansion under a certain substitution $\sigma$. The code verifying this
property is given below:

The entry point is depicted by the rule~\ref{rule:maybeeta} which takes a name
$n$, a term $t$ and a list of bound variables $L$ (originally it is the
singleton containing $n$). This rule checks if $t$ is a term of the form $T
\appsep n$ (for a term $T$), together with the auxiliary predicate
\elpiIn{reducible-to} which ensures if a term $t$ can reduce to a name $n$.
\marginpar{explain better: we start from $\lambda x.F x$}
The \elpiIn{maybe-eta} predicate dispatches the calls to \elpiIn{reducible-to};
three cases should be considered: 1) $t$ is a variable $v$, then $t$ can be an
$\eta$-expansion if at least one of the terms in the scope of $v$ is a
$\maybeeta$ of $n$; 2) $t$ is a lambda-term, then we recursively call
\elpiIn{maybe-eta} on the body of $t$ under a local name $x$ which is added to
the list $L$; 3) $t$ is an application, then $t$ is an $\eta$ expansion if i)
the last arguments of $t$ can be reduced one by one to the binders in the list
$L$ (we reverse the list in rule, since, by construction, this list is built in
reversed order) and ii) none of the first arguments of the application contain a
rigid occurrence of name in $L$.

As rapidly said before, \elpiIn{reducible-to} tells if a term $t$ reduce to a
name $n$, or equivalently if $\exists \sigma, \sigma t = n$. This predicate also
takes the list of all the binders explored (this list is originally empty). A
term $t$ reduces to a name $n$ if 1) $n = t$; 2) $t$ is a variable $v$, then $t$
reduce to a $n$ if it exists an arugment in the scope of $v$ reducing to $n$ and
forall name $n'$ in $L$, there is an argument reducing to $n'$ \marginpar{Note
that this function is partial: the arg in the scope should be distinct}; 3) $t$
is a lambda abstraction, then we call recursively \elpiIn{reducible-to} on the
body of the abstraction with a new local name added to the list $L$; 4) $t$ is
an application of $n$ to a list of arguments $L'$, then all the arguments should
reduce to the respective name in the list $L$.

Finally, a name $n$ occurs rigidly in a term $t$ if $n$ occurs in a subterm $t'$
of $t$ such that $t'$ does not appear in the scope of a variable.

An example of \maybeeta detection over the bound variable $x$ is the following:
\marginpar{maybe more examples in appendix}
%
\begin{gather}
  T = \lambda y. f \appsep A_{xy} \appsep (B \appsep a \appsep (\lambda z.y\appsep C_z)) \appsep D_x
\end{gather}

The correct call to \elpiIn{maybe-eta} is \elpiIn{maybe-eta x T L} with
\elpiIn{L = [x]}. At first we go under the abstraction $\lambda y$ adding $y$ to
$L$. Then we find an application, where we verify that 1) $A_{xy}$ does not
contain $x$ and $y$ rigidly which is the case; 2) $B \appsep a \appsep (\lambda z.y\appsep (C_z))$
and $D_x$ can respectively reduce to $y$ and $x$. The latter reduction is
evident, since $D$ has $x$ in scope; the former subterm can reduce to $y$ since
$B$ is a variable and it exists an argument ($\lambda z.y\appsep C_z$) reducible to
$y$: under the binder $z$, we have the application of $y$ with a variable with
$z$ in scope. Note that \maybeeta, only tells if it exists a substitution making
a term an eta expansion on any term $t$, i.e, $t$ can be in $\maybeeta \cup
\maybebeta$ without the constraint of being in \llambda, as our example shows. A
possible substitution making the term $T$ in the example an $\eta$-expansion is
$\sigma = \{A \mapsto \lambda x.\lambda y.a, B \mapsto \lambda x.\lambda y.y,
C \mapsto \lambda x.x, D \mapsto \lambda x.x\}$.

\paragraph{Compilation with \linketa} Thanks to the \elpiIn{maybe-eta}
predicate, we can detect ``$\eta$-problematic'' terms and, consequently replace
them with a fresh \Ho unification variable at compilation time. The code
below illustrate this dedicated compilation:
% TODO: maybe ref to \cref{sec:nutshell}

@@

\marginpar{Where This rule is put in the code ?}
This rule is applied on \Fo{} lambda-abstractions, tests if their body $F$ is an
\maybeeta wrt to a local fresh binder $x$, and if so, it compiles $F$ to the \Ho
term \elpiIn{F1} and returns a \Ho fresh variable $A$ having in scope the free
names occurring in \elpiIn{F1}. As sketch at the very end of
\cref{sec:lang-spec}, each time a subterm $t$ in \Fo is replaced with a \Ho
variable $v$, we build a \elpiIn{link-eta} between $v$ and and the term $t'$
obtained by the compilation of $t$.

A \elpiIn{link-eta}\footnote{\elpiIn{@val-link-eta A B} is syntactic sugar for
\elpiIn{val (link-eta A B)}}
is defined as

\noindent
We call the two terms carried by the link respectively left and right hand side,
called respectively \lhs and \rhs.

>>>>

\newtheorem{invariant}{Invariant}
\crefname{invariant}{invariant}{invariants}

@@

\begin{invariant}[Link left hand side]\label{inv:linklhs}
  The left hand side of a suspended link
  is a variable.
\end{invariant}

\noindent
New links are suspended by construction.
If the left hand side variable is assigned during a step, then 
the link is considered for progress and possibly eliminated.
This is discussed in \cref{sec:eta} and \cref{sc:beta}.

@@

Since compilation moves problematic terms out of the compiled terms,
\Ue{} can only perform a partial occur check. For example the
unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
but should be disallowed if a $\linkStore$ contains a link like
$\linketaM{}{Y}{\lambda z.X_z}$: We don't know yet if $Y$ will feature
a lambda in head position, but we surely know it contains $X$, hence
$f~Y$ and that triggers occur check.
The procedure \elpiIn{occur-check-links} is in charge of ensuring that
each link does not represent a (suspended) unification problem doomed
to fail because of occur check. This check is needed in order to
guarantee \cref{prop:fidelity} (simulation fidelity).

@@

\begin{invariant}[\linketa \rhs]
  The \rhs of a \linketa having the shape $\lambda x.F_x$
  where $F_x$ is a term not starting with the \elpiIn{lam} constructor.
  \label{inv:link-eta-right}
\end{invariant}

In the next three subsections we explain how we detect \maybeeta terms, how we
compile them and how \linketa are activated during the execution of the program
and provide justification for why \cref{inv:linklhs,inv:link-eta-right} remain true.

@@

TODO: clarify this
As a remark, note that $\lambda x. f \appsep A_x \appsep x$ is a \maybeeta,
since, despite, $x$ occurs in $f \appsep A_x$, it is still possible that this
subterm does not use $x$, for example if $A$ is a function on the form $\lambda
x.a$, where $a$ is a defined constant. In this case, the \maybeeta should
consider that the bound variable \elpiIn{x} does not ``rigidly'' occur in 
the given subterm.

@@

Thanks to the \elpiIn{maybe-eta} predicate, we can detect ``$\eta$-problematic''
terms and, consequently replace them with fresh \Ho unification variables at
compilation time. The code below illustrate how this relation is used to for
term compilation.

@@

This rule, to be inserted just before rule~\ref{rule:complam} from the code in
\cref{sec:compilation}, verifies if the \Fo{} term $t$ received in entry is a
\maybeeta. Let $\lambda t'$ be the compiled version of $t$, then the fresh
variable \elpiIn{A} returned as the new \Ho term as in scope all the free names
in $t'$. The critical part of this compilation is the creation of the \linketa,
which links the variable \elpiIn{A} with $t$. This link creation enforce
\cref{inv:linklhs} and \cref{inv:link-eta-right}, since clearly the \lhs is a
variable and the \rhs is a term starting with the \elpiIn{lam} constructor. We
can also note that for any name $x$, $tx = t' \appsep x$ never start with
\elpiIn{lam}, since by contradiction, if 1) $tx$ is an $\eta$-expansion, then
the recursive call to \elpiIn{comp} would have replaced $tx$ with a fresh
variable, which contradict the hypothesis and if 2) $t'$ is not an
$\eta$-expansion, then neither $\lambda t'$ is.
